Density-dependent selection and the limits of relative fitness

### 2018-03-16 manuscript.tex
l

There are a variety of different measures of fitness. Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is on relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures have different uses. But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection \citep{metcalf_2007,doebeli_2017}. While such a connection is fairly clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the ubiquitous use of relative fitness, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

@@

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we will use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. The situation is more complicated when we allow for population crowding. Since crowded and uncrowded conditions can favor quite different traits \citep{travis_2013}, the Malthusian parameters $r_i$ may respond differently to changes in $N$. The selection coefficient would then be density-dependent, and we would need to also know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot just independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise such a severe problem, because we only need to model the type frequencies, not also the ecological variable $N$.

In practice, many population genetics models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. Beyond being conceptually unsatisfactory \citep{doebeli_2017}, we might also doubt whether this density-independent-$s$, constant-$N$ setup is consistent with the ecology of many populations. Much of the seminal literature on density-dependent selection revolves around the connection between selection in crowded populations and the ability to keep growing at higher densities \citep{macarthur_1967,roughgarden_1979}. The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ (which will often entail a different equilibrium population density) excludes the others \citep{grover_1997}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. The Lotka-Volterra competition model also couples density to selection in crowded populations except in special cases \citep{smouse_1976,mallet_2012}. It would therefore seem that the ubiquitous density-independent-$s$, constant-$N$ setup is incompatible with a broad class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).

@@

In light of these difficulties, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}) and $N$ changes slowly \citep[Fig. 9.5]{otto_2011}. Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps.

However, our understanding of the strong-selection, unstable-$N$ regime is generally limited to a few prominent examples such as the classical density-dependent selection literature discussed above, and complex age-structured models \citep[Chap. 4]{charlesworth_1994}. Since constant relative fitness models are such a powerful idealization, it is important to understand the specifics of how and when they fail when selection is not weak, and $N$ is not stable. For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system? And if we make this switch, how much ecological detail do we need?

@@

We then outline the selection properties of our new variable-density lottery model (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
selection on the ability to contest territories, which behaves like a pure constant-$N$, constant-$s$ trait, and discuss how this fits with the classical notion of $K$-selection  (section ''The response of density to selection; $c$-selection versus $K$-selection''). We then consider selection on traits that regulate population density (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').

>>>>

There are a variety of different measures of fitness. Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}. In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is on relative genotypic proportions \cite[pp. 468]{barton_2007}. The variety of fitness measures is not problematic in itself, because different measures have different uses. But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp. 178]{charlesworth_1994}). While such a connection is fairly clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology. It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}. Given the ubiquitous use of relative fitness in evolutionary genetics, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.

@@

In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we will use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar). The selection coefficient $s$ is then simply the difference in intrinsic population growth rates. However, growth cannot continue at a constant rate indefinitely. As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g. \citealt[pp. 203]{begon_1990}). Since different traits are advantageous in crowded and uncrowded conditions, selection in the presence of crowding can be density-dependent, and we would then need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}. In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}. Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates. Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection only requires us to model the type frequencies, not also the ecological variable $N$.

In practice, many population genetics models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness. Selection is typically assumed to operate through viability (e.g. \citealt[pp. 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified. Beyond being conceptually unsatisfactory \citep{doebeli_2017}, we might also doubt whether this density-independent-$s$, constant-$N$ setup is consistent with the ecology of many populations. Much of the seminal literature on density-dependent selection revolves around the connection between selection in crowded populations and the ability to keep growing at higher densities \citep{macarthur_1967,roughgarden_1979}. The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a). Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ (which will often entail a different equilibrium population density) excludes the others \citep{grover_1997}. In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps. Density also changes in response to selection in the Lotka-Volterra competition model except in special cases \citep{smouse_1976,mallet_2012}. It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.

@@

In light of these difficulties, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp. 277]{ewens_2004}; \citealt[Chap. 4]{charlesworth_1994}). Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations. If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig. 9.5]{otto_2011}.

However, our understanding of the strong-selection, unstable-$N$ regime is

generally limited to a few prominent examples such as the classical density-dependent selection literature discussed above,

and complex age-structured models \citep[Chap. 4]{charlesworth_1994}.

Since constant relative fitness models are such a powerful idealization, it is important to understand the specifics of how and when they fail when selection is not weak, or $N$ is not stable. For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}. Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system? And if we make this switch, how much ecological detail do we need?

@@

We then outline the selection properties of our new variable-density lottery model (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
selection on the ability to contest territories, which behaves like a pure constant-$N$, constant-$s$ trait, and discuss how this fits with the classical notion of $K$-selection  (section ``The response of density to selection; $c$-selection versus $K$-selection''). We then consider selection on traits that regulate population density (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').
